Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand

### 2015-04-24 draftfinalreport_v3.Rnw
with lots maps (THIS IS THE ONE WE COULD KILL IF WE WANT THE MAIN MAPS GONE)

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks after measles introductions from abroad. In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region, and identify additional analyses that may help understand risk of infection in New Zealand. Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks from 2007 to 12~June~2014. We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of na\"{\i}ve people at the District Health Board (DHB) and national level. We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve required vaccination coverage levels. Our key findings are:

\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item The regression analyses suggest that age is a particularly strong risk factor for measles.  Increased wealth (NZDep 1--3) was a risk factor and people of European ethnicity make up the majority of cases. However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis. Pacific people are at greater risk per capita. However, this was driven by young Pacific children, as 2--24 year old Pacific people are less at risk than European and Maori, as are 5--17 year old Asian children.
\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 128,034 vaccinations nationally.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe. Peak overall overseas travel rates, and thus presumably risk from measles importation, is in December. However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include or exceed one, including 2013--2014 outbreaks. As analysed from data until 12 June 2014, the last outbreak to end in 2014 had an $R_v$ well above 1 (mean estimate 3.66, range 3.4--3.92). The range of estimates for $R_v$ between 2009--2014 was 0.18--3.92. These analyses suggest improved vaccination is a requisite to prevent measles persisting in the population and that large outbreaks could occur.
\item The cost of the first 187 measles cases in 2014 is estimated to be approximately \$1,041,186 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
\item The mean wage loss per measles case is estimated to be approximately \$839.
\item The mean cost of those measles case that attend hospital is estimated to be \$1,877 per attending case. Approximately 17\% of measles cases attend hospital.
\item The mean public health service cost per case is \$1,765 excluding hospitalisation costs.
\item The average number of contacts requiring quarantine was 2.11, requiring 7.3 days of quarantine on average, at a cost of \$170 per day.
\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed approximately \$5,365 per vaccine before the costs exceed the benefits for all DHBs.
\item Following catch up vaccination to ensure $R_v$ is < 1 measles introductions to New Zealand with a median size of 2 cases were predicted by simulation. Thus, typically individual cases and secondary cases will be the expectation. However, the mean outbreak size of 61 cases was predicted, because outbreak sizes of tens, hundreds, and very rarely up to many thousands of cases are predicted among the 1000 simulations following importation, despite $R_v$ being one and the outbreak predicted to die out. Thus, increased vaccination beyond the 28\% of the currently 11\% na\"{\i}ve population required may be useful to prevent these rare but costly events.
\end{itemize}

@@

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above). Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see \autoref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children, young adults and children under two years of age. Those under two years old are thought to be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

@@

We agree with the \emph {Progress Towards Measles Elimination in New Zealand - Final} report conclusions that testing for measles is performed appropriately within the required timeframe. Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007). However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onward, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.

@@

\begin{sidewaysfigure}
     %\centering
     \includegraphics[width=0.9\textwidth]{incidence_1997_2014.pdf}
     \caption{Measles notifications (confirmed and probable cases) in New Zealand from 1997 to 2014}
     \label{fig:incidence1997}
\end{sidewaysfigure}

@@

This large number of categories, some with small population sizes, leads to both overdispersion and zeroinflation, as there are many categories with zero cases, particularly in the adult age classes. Furthermore, initial preliminary multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions. Following discussion with the Ministry regarding the most appropriate NZDep categories, we reduced the number of categories from ten to three: NZDep 1--3 (least deprived), NZDep 4--7, and NZDep 8--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}). We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

@@

The numbers of cases per category and population sizes for the complete data set from 2007--2014 can be seen in \autoref{table:percap}. Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{table:percap}), leading to very large standard errors in regression analyses. However, there are also numerous issues with the data for MLA ethnicity category above and beyond the small population sizes, creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}). The breakdown of cases and per capita measles values can be see in \autoref{fig:cases_all} and \autoref{fig:percap_all}. Two way interactions are shown in the Appendix (\autoref{sec:app}, \autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).

\begin{table}[hbtp] \scriptsize
\captionsetup{width=1\textwidth, font={footnotesize}}
<<results=tex,echo=FALSE>>=
colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@
\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014. MLA is Middle Eastern/Latin American/African}
\label{table:percap}
\end{table}

For all our statistical analyses (including those above not shown) we used a Poisson error structure with rate adjusted for population size, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model. We used a model simplification approach, by beginning our analyses with all terms and all interactions, and then simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

\begin{align*} \label{eq:reg}
 y &\sim \textsf{Poisson}(\lambda n)\\
 \log(\lambda) &= \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _s (x_s) + \beta _a{}_e(x_a \times x_e)
  \end{align*}
Where $y$ is the number of cases, $\lambda$ is cases per capita, $n$ is population size, $\alpha$ is the intercept, $_a$ signifies age, $_e$ signifies Prioritized Ethnicity and $_s$ signifies NZDep.

@@

The cases and per capita measles rates are in \autoref{fig:cases_all} and \autoref{fig:percap_all}. The distribution of the measles cases per category used in the regression analyses are in \autoref{fig:histcasecat}. The predicted values from the regression model plotted against the reported cases are shown in \autoref{fig:predict}, and the residuals are shown in \autoref{fig:resid}. The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}). A summary of the regression model (\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in \autoref{table:regression_riskfactors}.

@@

Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case (\autoref{table:anova}). Indeed, all age categories are significantly less likely to be measles cases compared to 0--1 year olds, and the likelihood generally decreases with age (\autoref{table:regression_riskfactors}, \autoref{fig:cases_all}, \autoref{fig:percap_all}, see also \autoref{fig:ageinyears}).

@@

The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (\autoref{fig:ageandvac}). However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second. A further breakdown of those data suggest that majority of those `vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}). Plotting the unvaccinated cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).

@@

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{vacc_age.pdf}
\end{center}
\caption{Age of vaccination of vaccinated measles cases, 2007--12~June~2014. Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (4 years)}
\label{fig:vaccstat}
\end{figure}

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{dob_vacc_unvac.pdf}
\end{center}
\caption{Year of birth of unvaccinated of measles cases, 2007--12~June~2014}
\label{fig:yearandvac}
\end{figure}

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{dob_vacc_dose.pdf}
\end{center}
\caption{The year of birth for those vaccinated with measles, mumps, and rubella (MMR) dose 1 is shown in the upper panel, MMR dose 2 the lower panel.}
\label{fig:yearandvacf}
\end{figure}

@@

 \vspace{5mm} %5mm vertical space
\begin{table}
<<include=FALSE,echo=FALSE,results=tex>>=
latex(table_nir_mmr, file="", table.env=FALSE,rowname=NULL)
@
\caption{Summary of the proportion of 2006--2014 birth cohorts vaccinated against measles-mumps-rubella (MMR) dose 1 and MMR2 across area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12006}, \autoref{fig:fig22006}, \autoref{fig:fig12007}, \autoref{fig:fig22007}, \autoref{fig:fig12008}, \autoref{fig:fig22008}, \autoref{fig:fig12009}, \autoref{fig:fig22009}, \autoref{fig:fig12010}, \autoref{fig:fig22010}, \autoref{fig:fig12011}, \autoref{fig:fig22011}, \autoref{fig:fig12012}, \autoref{fig:fig22012}, \autoref{fig:fig12013}, \autoref{fig:fig22013}, \autoref{fig:fig12014}, and \autoref{fig:fig22014}.}
\label{table:tab_mmr_nir}
\end{table}

\begin{sidewaysfigure}
   \begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2006.pdf}
    \end{center}
    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR1 in each area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
%\end{center}
\label{fig:fig12006}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
\includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2006.pdf}
\end{center}
    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR2 in each area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig22006}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
\includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2007.pdf}
\end{center}
    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR1 in each area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig12007}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2007.pdf}
\end{center}
    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR2 in each area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig22007}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2008.pdf}
\end{center}
    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR1 in each area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig12008}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2008.pdf}
 \end{center}
    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR2 in each area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig22008}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2009.pdf}
 \end{center}
    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR1 in each area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
%\end{center}
\label{fig:fig12009}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2009.pdf}
 \end{center}
    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR2 in each area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
%\end{center}
\label{fig:fig22009}
\end{sidewaysfigure}

>>>>

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks after measles introductions from abroad. In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identify additional analyses that may help understand risk of infection in New Zealand. Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007 until 12 June 2014. We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of na\"{\i}ve people at the District Health Board (DHB) and national level. We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings are:

\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item The regression analyses suggest that age is a particularly strong risk factor for measles.  Increased wealth (NZDep 1--3) was a risk factor and people of European ethnicity make up the majority of cases. However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis. Pacific people are at greater risk per capita. However, this was driven by young Pacific children, as 2--24 year old Pacific people are less at risk than European and Maori, as are 5--17 year old Asian children.
\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 128,034 vaccinations nationally.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe. Peak overall overseas travel rates, and thus presumably risk from measles importation, is in December. However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include or exceed one, including 2013--2014 outbreaks. As analysed from data until 12 June 2014, the last outbreak to end in 2014 had an $R_v$ well above 1 (mean estimate 3.66, range 3.4--3.92). The range of estimates for $R_v$ between 2009--2014 was 0.18--3.92. These analyses suggest improved vaccination is a requisite to prevent measles persisting in the population and that large outbreaks could occur.
\item The cost of the first 187 measles cases in 2014 is estimated to be approximately \$1,041,186 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
\item The mean wage loss per measles case is estimated to be approximately \$839.
\item The mean cost per measles case attending hospital is estimated to be \$1,877 per case, and approximately 17\% of measles cases attend hospital
\item The mean public health service cost per case is \$1,765 excluding hospitalisation costs.
\item The average number of contacts requiring quarantine was 2.11, requiring 7.3 days of quarantine on average, at a cost of \$170.4 per day.
\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed approximately \$5,365 per vaccine before the costs exceed the benefits for all DHBs.
\item Following catch up vaccination to ensure $R_v$ is < 1 measles introductions to New Zealand with a median size of 2 cases were predicted by simulation. Thus, typically individual cases and secondary cases will be the expectation. However, the mean outbreak size of 61 cases was predicted, because outbreak sizes of tens, hundreds, and very rarely up to many thousands of cases are predicted among the 1000 simulations following importation, despite $R_v$ being one and the outbreak predicted to die out. Thus, increased vaccination beyond the 28\% of the currently 11\% na\'{\i}ve population required may be useful to prevent these rare but costly events.
\end{itemize}

@@

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above). Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see \autoref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children (especially young adults) and children under two years of age. Those under two year olds are thought to be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

@@

We agree with the \emph {Progress Towards Measles Elimination in New Zealand - Final} report conclusions that testing for measles is performed appropriately within the required timeframe. Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control. Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007). However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onward, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving. This is particularly in light of our modelling and risk analyses results (\autoref{sub:risk_analyses} and \autoref{section:epidemic_modelling}).

@@

\begin{sidewaysfigure}
     %\centering
     \includegraphics[width=0.9\textwidth]{incidence_1997_2014.pdf}
     \caption{Measles incidence in New Zealand from 1997 to 2014}
     \label{fig:incidence1997}
\end{sidewaysfigure}

@@

This large number of categories, some with small population sizes, leads to both overdispersion and zeroinflation, as there are many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to three: NZDep 1--3 (least deprived), NZDep 4--7, and NZDep 8--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}) following discussion with the Ministry regarding the most appropriate categories. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

@@

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}. Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{table:percap}), leading to very large standard error in regression analyses. However, there are also numerous issues with the data for MLA ethnicity category above and beyond the small population sizes, creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}). The breakdown of cases and per capita measles values can be see in \autoref{fig:cases_all} and \autoref{fig:percap_all}. Two way interactions are shown in the Appendix (\autoref{sec:app}, \autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).

\begin{table}[hbtp] \scriptsize
\captionsetup{width=1\textwidth, font={tiny}}
<<results=tex,echo=FALSE>>=
colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@
\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014. MLA is Middle Eastern/Latin American/African}
\label{table:percap}
\end{table}

For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model. We also account for differences in population sizes by using an offset term, the log(population size). We used a model simplification approach, by beginning our analyses with all terms and all interactions, and then simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

\begin{equation} \label{eq:reg}
 \log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a \times x_e)+ \log(population)  + \epsilon
  \end{equation}
Where $\alpha$ is the intercept, $y$ signifies cases, $_a$ signifies age, $_e$ signifies Prioritized Ethnicity, $_n$ signifies NZDep, and $\epsilon$ is the error term.

@@

The cases and per capita measles values are in \autoref{fig:cases_all} and \autoref{fig:percap_all}. The distribution of the measles cases per category used in the regression analyses are in \autoref{fig:histcasecat}. The predicted values from the regression model plotted against the reported cases are shown in \autoref{fig:predict}, and the residuals are shown in \autoref{fig:resid}. The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}). A summary of the regression model (\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in \autoref{table:regression_riskfactors}.

@@

Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case (\autoref{table:regression_riskfactors}). Indeed, all age categories are significantly less likely to be measles cases compared to 0--1 year olds, and the likelihood generally decreases with age (\autoref{table:regression_riskfactors}, \autoref{fig:cases_all}, \autoref{fig:percap_all}, see also \autoref{fig:ageinyears}).

@@

The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (\autoref{fig:ageandvac}). However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second. A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}). Plotting the unvaccinated cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).

@@

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{vacc_age.pdf}
\end{center}
\caption{Age of vaccination of vaccinated measles cases, 2007- 12 June 2014. Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (60 months)}
\label{fig:vaccstat}
\end{figure}

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{dob_vacc_unvac.pdf}
\end{center}
\caption{Year of birth of unvaccinated of measles cases, 2007- 12 June 2014}
\label{fig:yearandvac}
\end{figure}

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{dob_vacc_dose.pdf}
\end{center}
\caption{Year of birth of unvaccinated of measles cases, 2007- 12 June 2014. The year of birth for those vaccinated with measles, mumps, and rubella (MMR) dose 1 is shown in the upper panel, MMR dose 2 the lower panel.}
\label{fig:yearandvacf}
\end{figure}

@@

 \vspace{5mm} %5mm vertical space
\begin{table}
<<include=FALSE,echo=FALSE,results=tex>>=
latex(table_nir_mmr, file="", table.env=FALSE,rowname=NULL)
@
\caption{Summary of the proportion of 2006--2014 birth cohorts vaccinated against measles-mumps-rubella (MMR) dose 1 and MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12006}, \autoref{fig:fig22006}, \autoref{fig:fig12007}, \autoref{fig:fig22007}, \autoref{fig:fig12008}, \autoref{fig:fig22008}, \autoref{fig:fig12009}, \autoref{fig:fig22009}, \autoref{fig:fig12010}, \autoref{fig:fig22010}, \autoref{fig:fig12011}, \autoref{fig:fig22011}, \autoref{fig:fig12012}, \autoref{fig:fig22012}, \autoref{fig:fig12013}, \autoref{fig:fig22013}, \autoref{fig:fig12014}, and \autoref{fig:fig22014}.}
\label{table:tab_mmr_nir}
\end{table}

\begin{sidewaysfigure}
   \begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2006.pdf}
    \end{center}
    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR1 in each 2013 area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
%\end{center}
\label{fig:fig12006}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
\includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2006.pdf}
\end{center}
    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR2 in each 2013 area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig22006}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
\includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2007.pdf}
\end{center}
    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR1 in each 2013 area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig12007}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2007.pdf}
\end{center}
    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR2 in each 2013 area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig22007}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2008.pdf}
\end{center}
    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR1 in each 2013 area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig12008}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2008.pdf}
 \end{center}
    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR2 in each 2013 area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
% \end{center}
\label{fig:fig22008}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2009.pdf}
 \end{center}
    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR1 in each 2013 area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
%\end{center}
\label{fig:fig12009}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2009.pdf}
 \end{center}
    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR2 in each 2013 area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
%\end{center}
\label{fig:fig22009}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\begin{center}
    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2010.pdf}
\end{center}
    \caption{Proportion of the New Zealand population born in 2010 vaccinated against measles using MMR1 in each 2013 area unit. Note no adjustment is made for vaccine efficacy. The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange. See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
%\end{center}
\label{fig:fig12010}
\end{sidewaysfigure}
